Overlapping Batches for the Assessment of Solution Quality in Stochastic Programs

### 2013-10-11 omrp.tex
Rewrite large parts of the introduction
1. Include more motivation at the top, especially noting that we're
looking at assessing solution quality.
2. Include a list of contributions

We consider a stochastic program of the form
\begin{align} \tag{SP} \label{eq:sto_prog} 
	z^* & = \min_{\x \in X} \e{f(\x,\xit)},
\end{align}
where $\x$ is a vector of decision variables, $\xit$ is a vector of random variables, and $X$ is the feasible set, which is assumed to be independent of $\xit$.  
Further, it is assumed that the distribution of $\xit$ is known and that we can sample from it.  
Even though we will impose a more restrictive moment condition later, we assume that $\e{f(\x,\xit)}$ is well defined and finite for all $\x \in X$ and (\ref{eq:sto_prog}) has a finite optimal solution achieved on $X$.  
Let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (iid) sample from the distribution of $\xit$.  
A sampling approximation of (\ref{eq:sto_prog}) is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

We are interested in assessing the quality of a candidate solution $\xh \in X$, which may have been found in any way, defined by its optimality gap $\e{f(\xh,\xit)} - \zs$.  
This is important in practice since (\ref{eq:sto_prog}) typically cannot be solved exactly, so one only has an approximate solution $\xh$ without verification of its quality.  
Assessing solution quality is also a critical component of stopping criteria in algorithms.  
In many fields of optimization, it is common to evaluate an upper bound on the optimality gap through relaxations such as integrality or Lagrangian relations.  
In stochastic programming, given a sample size $m$, an upper bound on the optimality gap can be obtained by $\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}$ due to the inequality $\e{\zs_m} \leq \zs$.  
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$ \cite{Mak1999,norkin_pflug_ruszczynski_98}.  
For stochastic programs, one way to estimate an upper bound on optimality gaps is through Monte Carlo sampling. 
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, simply $\zs_m$ can be used.  
The resulting point estimator is  $G_m(\xh) = \frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m$.  
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms of $G_m(\xh)$.

@@

In this paper, we extend MRP by overlapping the batches similar to the overlapping batch means (OBM) introduced by Meketon and Schmeiser \cite{Meketon1984}. 
Suppose each batch contains $m$ observations and that there are a total of $2m$ observations.  
In the nonoverlapping case, there are only two batches: the first batch contains observations $\xit^1, \xit^2, \dots, \xit^m$ and the second batch contains observations $\xit^{m+1}, \xit^{m+2}, \dots, \xit^{2m}$.   
Meketon and Schmeiser \cite{Meketon1984} overlap the batches so that the first batch is as before, the second batch now consists of $\xit^2, \xit^3, \dots, \xit^{m+1}$, the third batch contains observations $\xit^3, \xit^4, \dots, \xit^{m+2}$ and so on. 
As a result, the corresponding batch means are no longer independent. 
However, by simply reusing the data in this fashion, and seemingly in a contradictory way that induces dependence, one obtains a better variance estimator: a variance estimator that asymptotically has two-thirds of the variance of the usual batch means variance estimator but the bias of the two estimators is approximately the same.  
The idea of overlapping batches can be used for variance estimators of non-means \cite{SAH90} and recently, has been extended to estimators based on standardized time series (e.g., area, Cram\'{e}r-von Mises) for steady-state simulation output analysis \cite{Alexopoulos01012007,Alexopoulos2007}. 
In this paper, we explore its use for assessing solution quality in stochastic programs.  
Similar to Welch \cite{Welch1987} and Song and Schmeiser \cite{Song1992}, we consider partial overlap between batches to reduce computational effort in our setting. 
We present conditions under which the point estimators are strongly consistent (that is, they converge almost surely (a.s.) as opposed in to in probability; referred to simply as consistent from this point on), the same variance reduction benefits are achieved, and the resulting confidence intervals are asymptotically valid.  
We present computational results on two-stage stochastic linear and integer programs with recourse to examine small sample behavior.  
Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.

>>>>

We wish to assess the quality of a candidate solution $\xh \in X$, which may have been found in any way, of a stochastic program of the form
\begin{align} \tag{SP} \label{eq:sto_prog} 
	z^* & = \min_{\x \in X} \e{f(\x,\xit)},
\end{align}
where $\x$ is a vector of decision variables, $\xit$ is a vector of random variables, and $X$ is the feasible set, which is assumed to be independent of $\xit$.
Further, it is assumed that the distribution of $\xit$ is known and that we can sample from it.
Even though we will impose a more restrictive moment condition later, we assume that $\e{f(\x,\xit)}$ is well defined and finite for all $\x \in X$ and (\ref{eq:sto_prog}) has a finite optimal solution achieved on $X$.

The quality of the candidate solution $\xh$ is defined by its optimality gap $\e{f(\xh,\xit)} - \zs$.
Assessing solution quality is important in practice since (\ref{eq:sto_prog}) typically cannot be solved exactly, so one only has an approximate solution $\xh$ without verification of its quality.
Assessing solution quality is also a critical component of stopping criteria in algorithms.
In many fields of optimization, it is common to evaluate an upper bound on the optimality gap through relaxations such as integrality or Lagrangian relations.
In stochastic programming, an approximation of \ref{eq:sto_prog} is often generated by sampling from the distribution of $\xi$.
Let $\xit^1, \xit^2, \dots, \xit^m$ be an independent and identically distributed (iid) sample from the distribution of $\xit$.
A sampling approximation of (\ref{eq:sto_prog}) is given by
\begin{align} \tag{SP$_m$} \label{eq:sto_prog_m}
	z_m^* & = \min_{\x \in X} \frac{1}{m} \sum_{i=1}^m f(\x,\xiti).
\end{align}
We denote an optimal solution to (\ref{eq:sto_prog}) as $\xs$ and an optimal solution to (\ref{eq:sto_prog_m}) as $\xs_m$.

This motivates estimating the upper bound on the optimality gap by using Monte Carlo sampling.
Given a sample size $m$, an upper bound on the optimality gap can be obtained by $\e{f(\xh,\xit)} - \zs \leq \e{f(\xh,\xit)} - \e{\zs_m}$ due to the inequality $\e{\zs_m} \leq \zs$.
This bound improves as the sample size increases, that is, $\e{\zs_m} \leq \e{\zs_{m+1}} \leq \zs$ \cite{Mak1999,norkin_pflug_ruszczynski_98}.  
A straightforward estimate of $\e{f(\xh,\xit)}$ is the sample mean, $\frac{1}{m} \sum_{i=1}^m f(\xh,\xiti)$. 
Instead of $\e{\zs_m}$, simply $\zs_m$ can be used.  
The resulting point estimator is  $G_m(\xh) = \frac{1}{m} \sum_{i=1}^m f(\xh,\xiti) - \zs_m$.  
Here, we assume the same observations $\xit^1, \dots, \xit^m$ are used in both terms of $G_m(\xh)$.

@@

In this paper, we extend MRP by overlapping the batches similar to the overlapping batch means (OBM) introduced by Meketon and Schmeiser \cite{Meketon1984}. 
As an example, suppose each batch contains $m$ observations and that there are a total of $2m$ observations.
In the nonoverlapping case, there are only two batches: the first batch contains observations $\xit^1, \xit^2, \dots, \xit^m$ and the second batch contains observations $\xit^{m+1}, \xit^{m+2}, \dots, \xit^{2m}$.   
Meketon and Schmeiser \cite{Meketon1984} overlap the batches so that the first batch is as before, the second batch now consists of $\xit^2, \xit^3, \dots, \xit^{m+1}$, the third batch contains observations $\xit^3, \xit^4, \dots, \xit^{m+2}$ and so on. 
As a result, the corresponding batch means are no longer independent. 
However, by simply reusing the data in this fashion, and seemingly in a contradictory way that induces dependence, one obtains a better variance estimator: a variance estimator that asymptotically has two-thirds of the variance of the usual batch means variance estimator but the bias of the two estimators is approximately the same.  
The idea of overlapping batches can be used for variance estimators of non-means \cite{SAH90} and recently, has been extended to estimators based on standardized time series (e.g., area, Cram\'{e}r-von Mises) for steady-state simulation output analysis \cite{Alexopoulos01012007,Alexopoulos2007}.

We explore the use of overlapping batches for assessing solution quality in stochastic programs.
Similar to Welch \cite{Welch1987} and Song and Schmeiser \cite{Song1992}, we consider partial overlap between batches to reduce computational effort in our setting.
The contributions of this paper are:
\begin{itemize}
	\item Formulate the overlapping multiple replications procedure for stochastic programs.
	\item Give conditions under which the point estimators are strongly consistent (that is, they converge almost surely (a.s.) as opposed in to in probability; referred to simply as consistent from this point on).
	\item Demonstrate that the classical variance reduction results are achieved for stochastic programs.
	\item Show the asymptotic validity of the resulting confidence intervals.
	\item Present computational results on two-stage stochastic linear and integer programs with recourse to examine small sample behavior.  
		Our experiments indicate that the asymptotic variance reduction is achieved with small sample sizes, while bias and coverage probability are unaffected.
\end{itemize}
